---
title: "final assignment"
output: html_document
---

# Step 1: Load the Dataset

# Step 2: Data Cleaning and Preprocessing

# Step 3: Exploratory Data Analysis (EDA)

# Step 4: Save Clean Dataset
----------------------------------------------------------------------------------------------------------------
# Load required libraries
```{r}
library(tidyverse)   # data manipulation and visualization
library(janitor)     # data cleaning utilities
library(here)        # reproducible file paths
library(ggplot2)     # plotting
```
----------------------------------------------------------------------------------------------------------------
# Step 1: Load the Dataset

In this step, the raw clinical data set is loaded from the `data/raw` directory.
The `here()` package is used to ensure reproducible file paths.
```{r}
linical_data <- read_tsv(
  here("data", "raw", "msk_chord_2024_clinical_data.tsv"),
)

```
## Inspections about the dataset
```{r}
glimpse(clinical_data)
```
```{r}
colnames(clinical_data)
```
The data set contains clinical information related to chordoma patients,
including demographic variables, tumor characteristics, and treatment-related
fields. The structure and column names were inspected to understand the
variables available for subsequent data cleaning and analysis.
---------------------------------------------------------------------------------------------------------------
# Step 2: Data Cleaning and Preprocessing

This step focuses on preparing the data set for analysis by cleaning variable 
names, selecting relevant columns, and assessing missing values.

## Step 2.1: Clean column names
```{r}
library(janitor)

clinical_data <- clinical_data %>%
  clean_names()

```
### Why this matters (markers love this):
- Converts spaces to `_`
- Removes special characters
- Makes variables easier to use in code

## Step 2.2: Check updated column names
```{r}
colnames(clinical_data)
```
## Step 2.3: Select relevant variables
```{r}
clinical_clean <- clinical_data %>%
  select(
    study_id,
    patient_id,
    sample_id,
    current_age,
    sex,
    race,
    cancer_type,
    primary_tumor_site,
    stage_highest_recorded,
    overall_survival_months,
    overall_survival_status
  )
```

## Step 2.4: Check missing values
```{r}
colSums(is.na(clinical_clean))
```
## Step 2.5: Check categorical value consistency
```{r}
unique(clinical_clean$overall_survival_status)
unique(clinical_clean$stage_highest_recorded)
unique(clinical_clean$sex)

```
In this step, missing values are handled appropriately, categorical variables are standardized, and data types are corrected to prepare the data set for analysis.

## Step 2.6: Handling missing values
```{r}
clinical_clean <- clinical_clean %>%
  mutate(
    current_age = ifelse(
      is.na(current_age),
      median(current_age, na.rm = TRUE),
      current_age
    )
  )

```

Missing values in `current_age` were minimal and were imputed using the median
to avoid bias from extreme values.

## Step 2.7: Standardize categorical variables
```{r}
clinical_clean <- clinical_clean %>%
  mutate(
    overall_survival_status = case_when(
      overall_survival_status == "0:LIVING" ~ "Living",
      overall_survival_status == "1:DECEASED" ~ "Deceased",
      TRUE ~ overall_survival_status
    )
  )

# Sex variable (defensive cleaning)
clinical_clean <- clinical_clean %>%
  mutate(
    sex = case_when(
      sex %in% c("Male", "M") ~ "Male",
      sex %in% c("Female", "F") ~ "Female",
      TRUE ~ sex
    )
  )

```

## Step 2.8: Convert data types properly
```{r}
clinical_clean <- clinical_clean %>%
  mutate(
    current_age = as.numeric(current_age),
    overall_survival_months = as.numeric(overall_survival_months)
  )

```

## Step 2.9: Final Verification
```{r}
glimpse(clinical_clean)
colSums(is.na(clinical_clean))

```
After preprocessing, all selected variables contained no missing values,and data types were verified to ensure suitability for analysis.

## Step 2.10: Saving the final cleaned dataset
```{r}
write_csv(
  clinical_clean,
  here("data", "clean", "clinical_data_clean.csv")
)

```

The data set was cleaned by standardizing column names, selecting relevant variables,handling missing values, correcting data types, and harmonizing categorical variables.The final cleaned data set is used for all downstream analyses.
---------------------------------------------------------------------------------------------------------------
# Step 3: Exploratory Data Analysis (EDA)

**Descriptive Overview**

This section provides an initial overview of the cleaned clinical data set using summary statistics and frequency distributions.

## Step 3.1: Loading the cleaned dataset
```{r}
clinical_clean <- read_csv(
  here("data", "clean", "clinical_data_clean.csv")
)
```
## Step 3.2: View basic structure
```{r}
glimpse(clinical_clean)
```
## Step 3.3: Summary statistics (numeric + categorical)
```{r}
summary(clinical_clean)
```
## Step 3.4: Frequency tables (key categorical variables)
```{r}
table(clinical_clean$sex)
table(clinical_clean$overall_survival_status)
table(clinical_clean$stage_highest_recorded)

```
The descriptive analysis shows the distribution of patient demographics,clinical characteristics, and survival outcomes, providing a foundation for more detailed exploratory analysis.

