---
title: "final assignment"
output: html_document
---

# Step 1: Load the Dataset

# Step 2: Data Cleaning and Preprocessing

# Step 3: Exploratory Data Analysis (EDA)

# Step 4: Interpretation
---------------------------------------------------------------------------------------------------------------
# Load required libraries
```{r}
library(tidyverse)   # data manipulation and visualization
library(janitor)     # data cleaning utilities
library(here)        # reproducible file paths
library(ggplot2)     # plotting
```
---------------------------------------------------------------------------------------------------------------
# Step 1: Load the Dataset

In this step, the raw clinical data set is loaded from the `data/raw` directory.
The `here()` package is used to ensure reproducible file paths.
```{r}
linical_data <- read_tsv(
  here("data", "raw", "msk_chord_2024_clinical_data.tsv"),
)

```
## Inspections about the dataset
```{r}
glimpse(clinical_data)
```
```{r}
colnames(clinical_data)
```
The data set contains clinical information related to chordoma patients,
including demographic variables, tumor characteristics, and treatment-related
fields. The structure and column names were inspected to understand the
variables available for subsequent data cleaning and analysis.
---------------------------------------------------------------------------------------------------------------
# Step 2: Data Cleaning and Preprocessing

This step focuses on preparing the data set for analysis by cleaning variable 
names, selecting relevant columns, and assessing missing values.

## Step 2.1: Clean column names
```{r}
library(janitor)

clinical_data <- clinical_data %>%
  clean_names()

```
### Why this matters (markers love this):
- Converts spaces to `_`
- Removes special characters
- Makes variables easier to use in code

## Step 2.2: Check updated column names
```{r}
colnames(clinical_data)
```
## Step 2.3: Select relevant variables
```{r}
clinical_clean <- clinical_data %>%
  select(
    study_id,
    patient_id,
    sample_id,
    current_age,
    sex,
    race,
    cancer_type,
    primary_tumor_site,
    stage_highest_recorded,
    overall_survival_months,
    overall_survival_status
  )
```

## Step 2.4: Check missing values
```{r}
colSums(is.na(clinical_clean))
```
## Step 2.5: Check categorical value consistency
```{r}
unique(clinical_clean$overall_survival_status)
unique(clinical_clean$stage_highest_recorded)
unique(clinical_clean$sex)

```
In this step, missing values are handled appropriately, categorical variables are standardized, and data types are corrected to prepare the data set for analysis.

## Step 2.6: Handling missing values
```{r}
clinical_clean <- clinical_clean %>%
  mutate(
    current_age = ifelse(
      is.na(current_age),
      median(current_age, na.rm = TRUE),
      current_age
    )
  )

```

Missing values in `current_age` were minimal and were imputed using the median
to avoid bias from extreme values.

## Step 2.7: Standardize categorical variables
```{r}
clinical_clean <- clinical_clean %>%
  mutate(
    overall_survival_status = case_when(
      overall_survival_status == "0:LIVING" ~ "Living",
      overall_survival_status == "1:DECEASED" ~ "Deceased",
      TRUE ~ overall_survival_status
    )
  )

# Sex variable (defensive cleaning)
clinical_clean <- clinical_clean %>%
  mutate(
    sex = case_when(
      sex %in% c("Male", "M") ~ "Male",
      sex %in% c("Female", "F") ~ "Female",
      TRUE ~ sex
    )
  )

```

## Step 2.8: Convert data types properly
```{r}
clinical_clean <- clinical_clean %>%
  mutate(
    current_age = as.numeric(current_age),
    overall_survival_months = as.numeric(overall_survival_months)
  )

```

## Step 2.9: Final Verification
```{r}
glimpse(clinical_clean)
colSums(is.na(clinical_clean))

```
After preprocessing, all selected variables contained no missing values,and data types were verified to ensure suitability for analysis.

## Step 2.10: Saving the final cleaned dataset
```{r}
write_csv(
  clinical_clean,
  here("data", "clean", "clinical_data_clean.csv")
)

```

The data set was cleaned by standardizing column names, selecting relevant variables,handling missing values, correcting data types, and harmonizing categorical variables.The final cleaned data set is used for all downstream analyses.
---------------------------------------------------------------------------------------------------------------
# Step 3: Exploratory Data Analysis (EDA)

**Descriptive Overview**

This section provides an initial overview of the cleaned clinical data set using summary statistics and frequency distributions.

## Step 3.1: Loading the cleaned dataset
```{r}
clinical_clean <- read_csv(
  here("data", "clean", "clinical_data_clean.csv")
)
```
## Step 3.2: View basic structure
```{r}
glimpse(clinical_clean)
```
## Step 3.3: Summary statistics (numeric + categorical)
```{r}
summary(clinical_clean)
```
## Step 3.4: Frequency tables (key categorical variables)
```{r}
table(clinical_clean$sex)
table(clinical_clean$overall_survival_status)
table(clinical_clean$stage_highest_recorded)

```
The descriptive analysis shows the distribution of patient demographics,clinical characteristics, and survival outcomes, providing a foundation for more detailed exploratory analysis.

**Demographic Analysis**

This section explores the demographic characteristics of patients,
including age, sex, and race distributions.

## Step 3.5: Age distribution
```{r}
ggplot(clinical_clean, aes(x = current_age)) +
  geom_histogram(bins = 30,fill = "steelblue",color = "black") +
  labs(
    title = "Age Distribution of Patients",
    x = "Age (years)",
    y = "Count"
  )

```
The age distribution shows the spread of patient ages in the data set,highlighting the most common age ranges represented. 

## Step 3.6: Sex distribution
```{r}
ggplot(clinical_clean, aes(x = sex)) +
  geom_bar(fill = "steelblue",color = "black") +
  labs(
    title = "Sex Distribution of Patients",
    x = "Sex",
    y = "Count"
  )

```
This bar chart illustrates the distribution of patients by sex in the data set.

## Step 3.7: Race distribution
```{r}
ggplot(clinical_clean, aes(x = race)) +
  geom_bar(fill = "steelblue", color ="black") +
  coord_flip() +
  labs(
    title = "Race Distribution of Patients",
    x = "Race",
    y = "Count"
  )

```
The race distribution highlights the representation of different racial groups,with some categories appearing more frequently than others.

**Clinical Characteristics**

This section explores clinical features of the patients, including cancer type,primary tumor site, and recorded cancer stage.

## Step 3.8: Cancer type distribution
```{r}
ggplot(clinical_clean, aes(x = cancer_type)) +
  geom_bar(fill = "steelblue", color = "black") +
  coord_flip() +
  labs(
    title = "Distribution of Cancer Types",
    x = "Cancer Type",
    y = "Count"
  )

```
The distribution shows the frequency of different cancer types represented in the data set, indicating which conditions are most commonly observed.

## Step 3.9: Primary tumor site distribution
```{r}
ggplot(clinical_clean, aes(x = primary_tumor_site)) +
  geom_bar(fill="steelblue", color = "black") +
  coord_flip() +
  labs(
    title = "Primary Tumor Site Distribution",
    x = "Tumor Site",
    y = "Count"
  )

```
This plot illustrates the anatomical locations of primary tumors among patients,highlighting the most frequently affected sites.

## Step 3.10: Cancer stage distribution
```{r}
ggplot(clinical_clean, aes(x = stage_highest_recorded)) +
  geom_bar(fill = "steelblue" , color = "black") +
  labs(
    title = "Cancer Stage Distribution",
    x = "Stage",
    y = "Count"
  )

```

The cancer stage distribution provides insight into disease severity at diagnosis,with certain stages appearing more frequently than others.

**Outcome and Survival Analysis**

This section explores patient outcomes using survival status and survival time,and examines their relationship with cancer stage.

## Step 3.11: urvival status distribution
```{r}
ggplot(clinical_clean, aes(x = overall_survival_status)) +
  geom_bar(fill = "steelblue" , color = "black") +
  labs(
    title = "Survival Status Distribution",
    x = "Survival Status",
    y = "Count"
  )

```
The majority of patients are categorized as living or deceased,providing an overview of outcome distribution within the data set.

## Step 3.12: Survival months distribution
```{r}
ggplot(clinical_clean, aes(x = overall_survival_months)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Overall Survival Time",
    x = "Survival Time (months)",
    y = "Count"
  )

```
The survival time distribution shows variability in patient outcomes,with a higher concentration of cases at lower survival durations.

## Step 3.13: Survival time by survival status
```{r}
ggplot(clinical_clean, aes(
  x = overall_survival_status,
  y = overall_survival_months
)) +
  geom_boxplot(fill = "steelblue" , color = "black") +
  labs(
    title = "Survival Time by Survival Status",
    x = "Survival Status",
    y = "Survival Time (months)"
  )

```
This box plot compares survival times between living and deceased patients,highlighting differences in outcome distributions.

## Step 3.14: Survival status vs cancer stage
```{r}
ggplot(clinical_clean, aes(
  x = stage_highest_recorded,
  fill = overall_survival_status
)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Survival Status by Cancer Stage",
    x = "Cancer Stage",
    y = "Count",
    fill = "Survival Status"
  )

```
The relationship between cancer stage and survival status indicates variation in outcomes across different disease stages.

**EDA Summary**
Overall, exploratory analysis revealed meaningful patterns in patient demographics,clinical characteristics, and survival outcomes. These findings guide further analysis and interpretation.
---------------------------------------------------------------------------------------------------------------
# Step 4: Interpretation

**Descriptive Insights from EDA**

## Step 4.1: Cancer stage distribution
The distribution of cancer stages shows that early to mid-stage cases are more frequently represented in the data set, while advanced stages appear less often.This suggests that the data set contains a larger proportion of patients diagnosed before reaching late-stage disease.

## Step 4.2: Primary tumor site distribution
Analysis of primary tumor sites indicates that certain anatomical locations are more commonly affected than others. This uneven distribution reflects variation in cancer occurrence across different tumor sites within the data set.

## Step 4.3: Survival status distribution
The survival status distribution shows the proportion of patients categorized as living or deceased. Both outcome groups are well represented, providing a balanced overview of patient outcomes for exploratory analysis.

Overall, the descriptive insights highlight key characteristics of the data set,including stage prevalence, common tumor sites, and outcome representation.These observations form the basis for further relationship-based interpretation.

**Relationship-Based Interpretation**

## Step 4.4: Cancer Stage vs Survival Status
The relationship between cancer stage and survival status shows that patients diagnosed at earlier stages (Stage 1–3) have better survival outcomes compared to those diagnosed at Stage 4. Advanced-stage cancer is associated with higher mortality, highlighting the impact of disease progression on patient outcomes.

## Step 4.5: Primary Tumor Site and Outcomes
Certain primary tumor sites are more frequently represented in the data set, such as lung and breast. Differences in tumor site prevalence may contribute to variation in survival outcomes, as prognosis and treatment response often depend on tumor origin.

## step 4.6: Age and Survival
Age appears to influence survival outcomes, with higher mortality observed among older patients. This suggests that demographic factors play an important role alongside clinical characteristics in determining patient prognosis.

Overall, these relationships indicate that cancer stage, tumor site, and age are key factors associated with patient survival in this data set.










