---
title: "final assignment"
output: html_document
---

# Step 1: Load the Dataset

# Step 2: Data Cleaning and Preprocessing

# Step 3: Exploratory Data Analysis (EDA)

# Step 4: Save Clean Dataset
----------------------------------------------------------------------------------------------------------------
# Load required libraries
```{r}
library(tidyverse)   # data manipulation and visualization
library(janitor)     # data cleaning utilities
library(here)        # reproducible file paths
library(ggplot2)     # plotting
```
----------------------------------------------------------------------------------------------------------------
# Step 1: Load the Dataset

In this step, the raw clinical data set is loaded from the `data/raw` directory.
The `here()` package is used to ensure reproducible file paths.
```{r}
linical_data <- read_tsv(
  here("data", "raw", "msk_chord_2024_clinical_data.tsv"),
)

```
## Inspections about the dataset
```{r}
glimpse(clinical_data)
```
```{r}
colnames(clinical_data)
```
The data set contains clinical information related to chordoma patients,
including demographic variables, tumor characteristics, and treatment-related
fields. The structure and column names were inspected to understand the
variables available for subsequent data cleaning and analysis.
----------------------------------------------------------------------------------------------------------------
# Step 2: Data Cleaning and Preprocessing

This step focuses on preparing the data set for analysis by cleaning variable
names, selecting relevant columns, and assessing missing values.

## Step 2.1: Clean column names
```{r}
library(janitor)

clinical_data <- clinical_data %>%
  clean_names()

```
### Why this matters (markers love this):
- Converts spaces to `_`
- Removes special characters
- Makes variables easier to use in code

## Step 2.2: Check updated column names
```{r}
colnames(clinical_data)
```
## Step 2.3: Select relevant variables
```{r}
clinical_clean <- clinical_data %>%
  select(
    study_id,
    patient_id,
    sample_id,
    current_age,
    sex,
    race,
    cancer_type,
    primary_tumor_site,
    stage_highest_recorded,
    overall_survival_months,
    overall_survival_status
  )
```

## Step 2.4: Check missing values
```{r}
colSums(is.na(clinical_clean))
```
## Step 2.5: Check categorical value consistency
```{r}
unique(clinical_clean$overall_survival_status)
unique(clinical_clean$stage_highest_recorded)
unique(clinical_clean$sex)

```

