---
title: "final assignment"
output: html_document
---

# Step 1: Load the Dataset

# Step 2: Data Cleaning and Preprocessing

# Step 3: Exploratory Data Analysis (EDA)

# Step 4: Save Clean Dataset
-----------------------------------------------------------------------------------------------
# Load required libraries
```{r}
library(tidyverse)   # data manipulation and visualization
library(janitor)     # data cleaning utilities
library(here)        # reproducible file paths
library(ggplot2)     # plotting
```

# Step 1: Load the Dataset

In this step, the raw clinical data set is loaded from the `data/raw` directory.
The `here()` package is used to ensure reproducible file paths.
```{r}
linical_data <- read_tsv(
  here("data", "raw", "msk_chord_2024_clinical_data.tsv"),
)

```
## Inspections about the dataset
```{r}
glimpse(clinical_data)
```
```{r}
colnames(clinical_data)
```
The data set contains clinical information related to chordoma patients,
including demographic variables, tumor characteristics, and treatment-related
fields. The structure and column names were inspected to understand the
variables available for subsequent data cleaning and analysis.
